---
title: "Annotation and Pathways"
author: "Tilo Alves Radtke 5444028"
date: "2026-07-10"
output: html_document
---



## R Markdown
```{r Working directory}
setwd("D:/Uni/Master/Data Science/Project/Code_local/MOFA")
```

MOFA Methylation Annotation + Enrichment Analysis Pipeline:
First we load the packages. The output is muted here, so set echo=TRUE if you encounter issues.
```{r setup, echo=FALSE}

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

packages <- c(
    "dplyr",
    "stringr",
    "tidyr",
    "readr",
    "clusterProfiler",
    "org.Hs.eg.db",
    "missMethyl",
    "IlluminaHumanMethylation450kanno.ilmn12.hg19"
)

for(pkg in packages){
    if(!requireNamespace(pkg, quietly = TRUE)){
        BiocManager::install(pkg, ask = TRUE)
    }
}

library(dplyr)
library(stringr)
library(tidyr)
library(readr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(missMethyl)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
```
Loading the top 500 features list we got from MOFA2, and extract the methylation data.
```{r}


INPUT_FILE <- "data/mofa_top_500_features_per_factor.csv"

TOP_N <- 500

OUTDIR <- "data/MOFA_Methylation_500_Analysis"

dir.create(OUTDIR, showWarnings = FALSE)


weights <- read.csv(
    INPUT_FILE,
    stringsAsFactors = FALSE
)


meth <- weights %>%
    filter(View == "Methylation") %>%
    mutate(
        CpG_ID = str_remove(Feature_Name, "_meth$")
    )

cat("Methylation probes:", nrow(meth), "\n")
```
We annotate the the 450 methylation data to the original hg19 human genome methylation reference using Illumina, and get Annotated_Methylation_Features.csv as the output.
```{r}

anno <- getAnnotation(
    IlluminaHumanMethylation450kanno.ilmn12.hg19
)

anno <- as.data.frame(anno)

anno$CpG_ID <- rownames(anno)

anno <- anno %>%
    dplyr::select(
        CpG_ID,
        chr,
        pos,
        strand,
        UCSC_RefGene_Name,
        UCSC_RefGene_Group,
        Relation_to_Island
    )


meth <- meth %>%
    left_join(anno, by = "CpG_ID")
#summary(meth)

clean_field <- function(x){

    if(is.na(x) || x == "")
        return("")

    vals <- unique(unlist(str_split(x, ";")))

    vals <- vals[vals != ""]

    paste(vals, collapse = ";")
}

meth <- meth %>%
    rowwise() %>%
    mutate(
        Gene = clean_field(UCSC_RefGene_Name),
        Gene_Region = clean_field(UCSC_RefGene_Group)
    ) %>%
    ungroup()


write.csv(
    meth,
    file.path(OUTDIR,
              "Annotated_Methylation_Features.csv"),
    row.names = FALSE
)
```


Here we analyse the significance of the CPGs, Positive / Negative Loadings, missMethyl GO enrichment and Gene-based analyses using GO enrichment and KEGG per factor. For our results, we especially used the missMethyl GO enrichment results, and decided based on its FDR, whether a pathway is significantly present in the factor or not on a FDR<0.05 cutoff.
WARNING: This part takes a while to calculate, not hours but multiple minutes.
```{r}

all_cpg <- unique(meth$CpG_ID)


factors <- unique(meth$Factor)

for(f in factors){

    cat("Calculation of significance of the annotation:", f, "\n")


    factor_dir <- file.path(OUTDIR, f)

    dir.create(factor_dir,
               recursive = TRUE,
               showWarnings = FALSE)

    
    # Top CpGs
    

    top_cpgs <- meth %>%
        filter(Factor == f) %>%
        arrange(desc(Absolute_Weight)) %>%
        slice_head(n = TOP_N)

    write.csv(
        top_cpgs,
        file.path(
            factor_dir,
            paste0(f, "_TopCpGs.csv")
        ),
        row.names = FALSE
    )

    #We get a lot of files, so we gather the important part into a summary:

    island_summary <- as.data.frame(
        table(top_cpgs$Relation_to_Island)
    )

    write.csv(
        island_summary,
        file.path(
            factor_dir,
            paste0(f,
                   "_CpG_Island_Distribution.csv")
        ),
        row.names = FALSE
    )

    region_summary <- as.data.frame(
        table(top_cpgs$Gene_Region)
    )

    write.csv(
        region_summary,
        file.path(
            factor_dir,
            paste0(f,
                   "_GeneRegion_Distribution.csv")
        ),
        row.names = FALSE
    )

    #Separate Positive / Negative Loadings

    positive <- top_cpgs %>%
        filter(Raw_Weight > 0)

    negative <- top_cpgs %>%
        filter(Raw_Weight < 0)

    write.csv(
        positive,
        file.path(
            factor_dir,
            paste0(f, "_PositiveCpGs.csv")
        ),
        row.names = FALSE
    )

    write.csv(
        negative,
        file.path(
            factor_dir,
            paste0(f, "_NegativeCpGs.csv")
        ),
        row.names = FALSE
    )


    #missMethyl GO enrichment
 

    sig_cpg <- unique(top_cpgs$CpG_ID)

    if(length(sig_cpg) >= 10){

        cat("Running missMethyl GO...\n")

        go_results <- gometh(
            sig.cpg = sig_cpg,
            all.cpg = all_cpg,
            collection = "GO"
        )

        write.csv(
            go_results,
            file.path(
                factor_dir,
                paste0(f,
                       "_missMethyl_GO.csv")
            ),
            row.names = FALSE
        )
    }


    #Gene-based analyses


    genes <- unique(
        unlist(
            str_split(
                paste(top_cpgs$Gene,
                      collapse = ";"),
                ";"
            )
        )
    )

    genes <- genes[
        genes != "" &
        !is.na(genes)
    ]

    cat("Genes:", length(genes), "\n")

    if(length(genes) >= 10){

        # GO Enrichment
        
        ego <- enrichGO(
            gene = genes,
            OrgDb = org.Hs.eg.db,
            keyType = "SYMBOL",
            ont = "BP",
            pAdjustMethod = "BH"
        )

        if(!is.null(ego)){

            ego_df <- as.data.frame(ego)

            write.csv(
                ego_df,
                file.path(
                    factor_dir,
                    paste0(f,
                           "_GO_BP.csv")
                ),
                row.names = FALSE
            )
        }

        # KEGG

        gene_map <- tryCatch({

            bitr(
                genes,
                fromType = "SYMBOL",
                toType = "ENTREZID",
                OrgDb = org.Hs.eg.db
            )

        }, error = function(e) NULL)

        if(!is.null(gene_map)){

            kegg <- enrichKEGG(
                gene = gene_map$ENTREZID,
                organism = "hsa"
            )

            if(!is.null(kegg)){

                kegg_df <- as.data.frame(kegg)

                write.csv(
                    kegg_df,
                    file.path(
                        factor_dir,
                        paste0(f,
                               "_KEGG.csv")
                    ),
                    row.names = FALSE
                )
            }
        }
    }
}
```
Now a Summary of all factors is written to data/MOFA_Methylation_500_Analysis/Factor_Summary.csv
```{r}

summary_table <- meth %>%
    group_by(Factor) %>%
    summarise(
        N_CpGs = n(),
        Mean_Weight = mean(Absolute_Weight),
        Median_Weight = median(Absolute_Weight),
        .groups = "drop"
    )

write.csv(
    summary_table,
    file.path(
        OUTDIR,
        "Factor_Summary.csv"
    ),
    row.names = FALSE
)

cat("\nAnalysis complete.\n")
cat("Results written to:", OUTDIR, "\n")
```


Due to the amount of results, we directly filter by the FDR<0.05(calculated by gometh()), and get out missmethyl values to the file Significant_missMethyl_GO_FDR_0_05.csv, which is the final datatable from the methylation data:
```{r}
all_sig_go <- list()

for(f in factors){

    go_file <- file.path(
        OUTDIR,
        f,
        paste0(f, "_missMethyl_GO.csv")
    )

    if(file.exists(go_file)){

        go_res <- read.csv(
            go_file,
            stringsAsFactors = FALSE
        )

        sig_res <- go_res[
            go_res$FDR < 0.05,
        ]

        if(nrow(sig_res) > 0){

            sig_res$Factor <- f

            all_sig_go[[f]] <- sig_res
        }
    }
}

if(length(all_sig_go) > 0){

    significant_GO <- do.call(
        rbind,
        all_sig_go
    )

    # Put Factor as first column
    significant_GO <- significant_GO[
        ,
        c(
            "Factor",
            setdiff(
                colnames(significant_GO),
                "Factor"
            )
        )
    ]

    write.csv(
        significant_GO,
        file.path(
            OUTDIR,
            "Significant_missMethyl_GO_FDR_0_05.csv"
        ),
        row.names = FALSE
    )

    cat(
        "\nFound",
        nrow(significant_GO),
        "significant GO terms.\n"
    )

} else {

    cat(
        "\nNo missMethyl GO terms with FDR < 0.05 found.\n"
    )
}

```
In this last codeblock, it creates a bubbleplot of the significant pathways per factor. We didnt use it in the presentation due to being not well readable.
```{r}
library(dplyr)
library(ggplot2)


meth <- read.csv(
    "data/MOFA_Methylation_500_Analysis/Significant_missMethyl_GO_FDR_0_05.csv",
    stringsAsFactors = FALSE
)

# ============================================================
# SCORE
# ============================================================

meth <- meth %>%
    mutate(
        Score = -log10(FDR)
    )

# ============================================================
# KEEP TOP 10 TERMS PER FACTOR
# ============================================================

meth_plot <- meth %>%
    group_by(Factor) %>%
    arrange(FDR) %>%
    slice_head(n = 10) %>%
    ungroup()

# ============================================================
# BUBBLEPLOT
# ============================================================

p <- ggplot(
    meth_plot,
    aes(
        x = Factor,
        y = TERM
    )
) +
    geom_point(
        aes(
            size = DE,
            color = Score
        )
    ) +
    theme_bw() +
    theme(
        axis.text.y = element_text(size = 8)
    ) +
    labs(
        title = "Methylation GO Enrichment",
        size = "Differentially represented genes",
        color = "-log10(FDR)"
    )

ggsave(
    "data/MOFA_Methylation_500_Analysis/Methylation_GO_Bubbleplot.pdf",
    p,
    width = 12,
    height = 10
)

```
