---
title: "Plotting_annotations"
author: "Tilo Alves Radtke 5444028"
date: "2026-07-12"
output: html_document
---
```{r working directory}
# Working directory: this file expects to run from the MOFA/ folder.
# knitr sets this automatically when knitting. Uncomment and edit only if you
# run the chunks interactively from a different location.
# setwd("path/to/MOFA")
```


MOFA RNA Annotation + Enrichment Analysis Pipeline:
First we load the packages. The output is muted here, so set echo=TRUE if you encounter issues.
```{r setup, echo=FALSE}


if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

pkgs <- c(
    "dplyr",
    "clusterProfiler",
    "org.Hs.eg.db",
    "AnnotationDbi",
    "ggplot2"
)

for(pkg in pkgs){
    if(!requireNamespace(pkg, quietly = TRUE)){
        BiocManager::install(pkg, ask = TRUE)
    }
}

library(dplyr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(ggplot2)

```

We first load the data. as input the top 500 features of each factor is used.
```{r loading}
#INPUT_FILE <- "data/feature_weights.csv"

INPUT_FILE <- "data/mofa_top_500_features_per_factor.csv"
TOP_N <- 500

OUTDIR <- "data/MOFA_RNA_Anno"

dir.create(
    OUTDIR,
    showWarnings = FALSE
)

weights <- read.csv(
    INPUT_FILE,
    stringsAsFactors = FALSE
)


rna <- weights %>%
    dplyr::filter(
        grepl(
            "RNA|rna|expression",
            View,
            ignore.case = TRUE
        )
    )

cat(
    "RNA features found:",
    nrow(rna),
    "\n"
)

#This just removes the suffix we attached in MOFA
rna$Gene <- gsub(
    "_rna$",
    "",
    rna$Feature_Name
)

```

We annotate the RNA features using org.Hs.eg.db. This uses the human annotation database.
```{r annotation}
anno <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys = unique(rna$Gene),
    keytype = "SYMBOL",
    columns = c(
        "SYMBOL",
        "ENTREZID",
        "GENENAME"
    )
)

rna <- dplyr::left_join(
    rna,
    anno,
    by = c(
        "Gene" = "SYMBOL"
    )
)

rna <- rna %>%
    dplyr::filter(
        !is.na(ENTREZID)
    )

write.csv(
    rna,
    file.path(
        OUTDIR,
        "Annotated_RNA_Features.csv"
    ),
    row.names = FALSE
)

factors <- unique(
    rna$Factor
)

all_results <- list()

```

Here we use enrichGO() to perform a Gene Ontology enrichment on the feature genes. It also calculates, whether the results are significant with a FDR < 0.05.
```{r GO ENRICHMENT}
for(f in factors){

    cat(
        "\nProcessing",
        f,
        "\n"
    )

    factor_dir <- file.path(
        OUTDIR,
        f
    )

    dir.create(
        factor_dir,
        recursive = TRUE,
        showWarnings = FALSE
    )

    tmp <- rna %>%
        dplyr::filter(
            Factor == f
        )

    if(nrow(tmp) < 20){

        cat(
            "Skipping:",
            f,
            "\n"
        )

        next
    }

    #goes over the top features

    tmp <- tmp %>%
        dplyr::arrange(
            desc(abs(Raw_Weight))
        ) %>%
        dplyr::slice_head(
            n = TOP_N
        )

    write.csv(
        tmp,
        file.path(
            factor_dir,
            paste0(
                f,
                "_Top500Genes.csv"
            )
        ),
        row.names = FALSE
    )

    genes <- unique(
        tmp$ENTREZID
    )

    genes <- genes[
        !is.na(genes)
    ]

    if(length(genes) < 10){

        next
    }

    # the actual GO enrichment

    ego <- tryCatch({

        enrichGO(
            gene = genes,
            OrgDb = org.Hs.eg.db,
            keyType = "ENTREZID",
            ont = "BP",
            pAdjustMethod = "BH",
            readable = TRUE
        )

    }, error = function(e){

        NULL

    })

    if(is.null(ego))
        next

    go_df <- as.data.frame(
        ego
    )

    if(nrow(go_df) == 0)
        next

    go_df$Factor <- f

    write.csv(
        go_df,
        file.path(
            factor_dir,
            paste0(
                f,
                "_GO.csv"
            )
        ),
        row.names = FALSE
    )

    sig_go <- go_df %>%
        dplyr::filter(
            p.adjust < 0.05
        )

    write.csv(
        sig_go,
        file.path(
            factor_dir,
            paste0(
                f,
                "_GO_Significant.csv"
            )
        ),
        row.names = FALSE
    )

    if(nrow(sig_go) > 0){

        all_results[[f]] <- sig_go

    }
}
```


All significant results are put into the table All_Factors_Significant_GO.csv for all factors.
```{r Sign_results}
if(length(all_results) == 0){

    stop(
        "No significant pathways found."
    )
}

combined <- dplyr::bind_rows(
    all_results
)

write.csv(
    combined,
    file.path(
        OUTDIR,
        "All_Factors_Significant_GO.csv"
    ),
    row.names = FALSE
)

#KEEP TOP 10 TERMS PER FACTOR

plot_df <- combined %>%
    dplyr::group_by(
        Factor
    ) %>%
    dplyr::arrange(
        p.adjust
    ) %>%
    dplyr::slice_head(
        n = 10
    ) %>%
    dplyr::ungroup()

plot_df$Score <- -log10(
    plot_df$p.adjust
)


```

This codeblock creates a unused Bubbleplot of the significant pathway results per factor.
```{r Bubbleplot}
p <- ggplot(
    plot_df,
    aes(
        x = Factor,
        y = Description
    )
) +

geom_point(
    aes(
        size = Count,
        color = Score
    )
) +

theme_bw() +

theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  # slant x-axis labels
) +

labs(
    title = "RNA GO Enrichment (Top 500 Genes per Factor)",
    size = "Gene Count",
    color = "-log10(FDR)"
)

ggsave(
    file.path(
        OUTDIR,
        "RNA_GO_Bubbleplot.pdf"
    ),
    p,
    width = 12,
    height = 10
)
```

The top 10 pathways per factor are put into Top_Pathway_Per_Factor.csv, to create a overview of what the different factors do. There are pathways, that were cut out of factor 1 and 2, that were more significant than any pathways from factor 14 f.E., but this is about getting a overview of all factors in "Top annotation plots RNA_METH.Rmd".
```{r}
summary_df <- plot_df %>%
    dplyr::group_by(
        Factor
    ) %>%
    dplyr::slice_min(
        p.adjust,
        n = 1
    ) %>%
    dplyr::ungroup()

write.csv(
    summary_df,
    file.path(
        OUTDIR,
        "Top_Pathway_Per_Factor.csv"
    ),
    row.names = FALSE
)

cat(
    "\nFinished.\n"
)

cat(
    "\nResults written to:",
    OUTDIR,
    "\n"
)
```


