---
title: "TCGA - KIRC Cohort Data Download: Kidney Clear Cell Carcinoma Downloading Data Files"
author: 'Group 7: Maksim Danilchyk, Sofya Shorzhina, Tilo Alves Radtke, Kristian Reinhart'
#date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: '3'
    latex_engine: xelatex
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  cache     = FALSE,
  fig.align = "center"
)
```

------------------------------------------------------------------------

# Overview

This script serves to download the TCGA KIRC Cohort data files for a consistent folder structure and file naming.

The TCGA KIRC Dataset can be found on the Xenabrowser webpages hosted by
the University of Santa Cruz:
[https://xenabrowser.net/datapages/?cohort=TCGA%20Kidney%20Clear%20Cell%20Carcinoma%20(KIRC)](https://xenabrowser.net/datapages/?cohort=TCGA%20Kidney%20Clear%20Cell%20Carcinoma%20(KIRC)){.uri}

The dataset offers several different modalities, most notably the
following ones:

- RNAseq gene expression, whole genome, IlluminaHiSeq_RNASeqV2
  - log2(norm_count+1) transformed, 20.531 identifiers X 606 samples
  - Download Link:
    <https://xenabrowser.net/datapages/?dataset=TCGA.KIRC.sampleMap%252FHiSeqV2>
  - Original Filename: HiSeqV2.gz
  - Renamed Filename: HiSeqV2.txt
- Exon expression RNAseq, IlluminaHiSeq_RNASeqV2
  - log2(norm_count+1) transformed, 239.323 identifiers X 606 samples
  - Download Link:
    <https://xenabrowser.net/datapages/?dataset=TCGA.KIRC.sampleMap%252FHiSeqV2_exon>
  - Original Filename: HiSeqV2_exon.gz
  - Renamed Filename: HiSeqV2_exon.txt
- DNA methylation, Illumina Infinium HumanMethylation450
  - beta value transformed, 485.578 identifiers X 480 samples
  - Download Link:
    <https://xenabrowser.net/datapages/?dataset=TCGA.KIRC.sampleMap%252FHumanMethylation450>
  - Original Filename: HumanMethylation450.gz
  - Renamed Filename: Methylation450.txt
- Phenotypes
  - Clinical metadata such as patient age, sex, diagnosis, etc.; 945
    samples X 112 identifiers
  - Download Link:
    <https://xenabrowser.net/datapages/?dataset=TCGA.KIRC.sampleMap%252FKIRC_clinicalMatrix>
  - Original Filename: KIRC_clinicalMatrix
  - Renamed Filename: KIRC_clinicalMatrix.txt
- Curated Survival Data
  - Curated survival data including OS: overall survial, PFI:
    progression-free interval, DSS: disease-specific survival, DFI:
    disease-free interval; 944 samples X 11 identifiers
  - Download Link:
    <https://xenabrowser.net/datapages/?dataset=survival%252FKIRC_survival.txt>
  - Original Filename: KIRC_survival.txt
  - Renamed Filename: KIRC_survival.txt
- Somatic Mutation (SNP and INDEL) - MC3 gene-level non-silent mutation
  - binary non-silent mutation call, 40.544 identifiers X 368 samples
  - Download Link:
    <https://xenabrowser.net/datapages/?dataset=mc3_gene_level%252FKIRC_mc3_gene_level.txt>
  - Original Filename: KIRC_mc3_gene_level.txt.gz
  - Renamed Filename: KIRC_mc3_gene_level.txt
- Thresholded gene-level Copy Number Variation (CNV), GISTIC2 method
  - Copy number (gene-level) thresholded to integer values, 24.777
    identifiers X 528 samples
  - Download Link:
    <https://xenabrowser.net/datapages/?dataset=TCGA.KIRC.sampleMap%252FGistic2_CopyNumber_Gistic2_all_thresholded.by_genes>
  - Original Filename:
    Gistic2_CopyNumber_Gistic2_all_thresholded.by_genes.gz
  - Renamed Filename: Gistic2_CopyNumber_Gistic2_all_thresholded.by_genes.txt
- Protein expression RPPA
  - Protein expression by reverse phase protein array (RPPA), normalized RPPA value, 274 identifiers X 478 samples
  - Download Link:
    <https://xenabrowser.net/datapages/?dataset=TCGA.KIRC.sampleMap%252FRPPA>
  - Original Filename: RPPA.gz
  - Renamed Filename: RPPA.txt

Additionally of interest are the codes for encoding the sample type
(i.e. the type of the sample such as "primary tumor", "normal tissue",
etc.) whose explanation can be found here:\
<https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/sample-type-codes>

As well as a more general overview of the sample code meaning:\
<https://docs.gdc.cancer.gov/Encyclopedia/pages/TCGA_Barcode/>


```{r Loading Libraries}
# Check/install required package
if (!requireNamespace("R.utils", quietly = TRUE)) {
    install.packages("R.utils")
}
```


```{r Download Helper Function}
getDataSet <- function(downloadLink,filename,newFilename,downloadDirectory,doUnzip,timeOut=600){
  # Function to download, extract and rename the necessary dataset files
  # downloadLink: Link to the file to be donwloaded
  # filename: Filename of the donwloaded file / archive, including extension. Leave empty ("") if no renaming necessary
  # newFilename: New filename for the downloaded file. File will be renamed to this. If doUnzip=TRUE renaming is performed on the extracted file.
  # downloadDirectory: Directory where to download the file to
  # doUnzip: Logical. TRUE if the downloaded file is an archive to be extracted
  # timeout: Timeout in seconds for the download. Default is 600 = 10 minutes, should be plenty for the given files
  tryCatch({
    # Create directory if it doesn't exist
    if (!dir.exists(downloadDirectory)) {
      dir.create(downloadDirectory, recursive = TRUE)
    }
    
    # Define original download filename including filepath
    downloadPath <- file.path(downloadDirectory, filename)
    # Define renamed filename including filepath
    finalPath <- file.path(downloadDirectory, newFilename)
    
    options(timeout = timeOut) # Give a download timeout of 10 minutes for the larger files
    # Download file
    status <- download.file(
      downloadLink,
      destfile = downloadPath,
      mode = "wb",
      quiet = TRUE
    )
    
    # Check if download failed, abort if it did
    if (status!=0) {
      stop("Download failed.")
    }
    
    # Extract and rename the file if needed
    if (doUnzip) {
      R.utils::gunzip(
        filename = downloadPath,
        destname = finalPath,
        remove = FALSE,
        overwrite = TRUE
      )
    } else {
      # Rename the downloaded file (if no unzipping necessary)
      if (newFilename!="" && !file.rename(downloadPath, finalPath)) {
        # Abort if renaming fails for some reason
        if(newFilename!="") {
            # Only give warning if need for renaming exists in the first place
            stop(paste0("Failed to rename downloaded file. Does the file '",downloadPath,"' exist?"))
        }
      }
    }
    
    # Delete original compressed / downloaded file
    if (file.exists(downloadPath) && newFilename!="" && !file.remove(downloadPath)) {
      warning(paste0("Downloaded file could not be removed. Does the file '",downloadPath,"' exist?"))
    }
  }, error = function(e) {
    # Something unexpected went wrong
    warning("Downloading / extracting / renaming the dataset from '",downloadLink,"' went wrong somewhere.")
    message("getDataSet failed: ", conditionMessage(e))
  })
}
```

```{r, Downloading Data}
# Download, extract and rename the KIRC RNASeq File
getDataSet(
    downloadLink="https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/TCGA.KIRC.sampleMap%2FHiSeqV2.gz",
    filename="HiSeqV2.gz",
    newFilename="HiSeqV2.txt",
    downloadDirectory="TCGA-KIRC",
    doUnzip=TRUE)

# Download, extract and rename the KIRC Methylation 450k File
getDataSet(
    downloadLink="https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/TCGA.KIRC.sampleMap%2FHumanMethylation450.gz",
    filename="HumanMethylation450.gz",
    newFilename="HumanMethylation450.txt",
    downloadDirectory="TCGA-KIRC",
    doUnzip=TRUE)

# Download, extract and rename the KIRC Gene Level Copy Number (Thresholded) File
getDataSet(
    downloadLink="https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/TCGA.KIRC.sampleMap%2FGistic2_CopyNumber_Gistic2_all_thresholded.by_genes.gz",
    filename="Gistic2_CopyNumber_Gistic2_all_thresholded.by_genes.gz",
    newFilename="Gistic2_CopyNumber_Gistic2_all_thresholded.by_genes.txt",
    downloadDirectory="TCGA-KIRC",
    doUnzip=TRUE)

# Download, extract and rename the KIRC Gene Level Copy Number (non-thresholded) File
getDataSet(
    downloadLink="https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/TCGA.KIRC.sampleMap%2FGistic2_CopyNumber_Gistic2_all_data_by_genes.gz",
    filename="Gistic2_CopyNumber_Gistic2_all_data_by_genes.gz",
    newFilename="Gistic2_CopyNumber_Gistic2_all_data_by_genes.txt",
    downloadDirectory="TCGA-KIRC",
    doUnzip=TRUE)

# Download, extract and rename the KIRC Somatic mutations File
getDataSet(
    downloadLink="https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/mc3_gene_level%2FKIRC_mc3_gene_level.txt.gz",
    filename="KIRC_mc3_gene_level.txt.gz",
    newFilename="KIRC_mc3_gene_level.txt",
    downloadDirectory="TCGA-KIRC",
    doUnzip=TRUE)

# Download, extract and rename the KIRC Exon expression over whole genome File
getDataSet(
    downloadLink="https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/TCGA.KIRC.sampleMap%2FHiSeqV2_exon.gz",
    filename="HiSeqV2_exon.gz",
    newFilename="HiSeqV2_exon.txt",
    downloadDirectory="TCGA-KIRC",
    doUnzip=TRUE)

# Download, extract and rename the KIRC Protein Expression RPPA File
getDataSet(
    downloadLink="https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/TCGA.KIRC.sampleMap%2FRPPA.gz",
    filename="RPPA.gz",
    newFilename="RPPA.txt",
    downloadDirectory="TCGA-KIRC",
    doUnzip=TRUE)

# Download, extract and rename the KIRC Phenotypes File
getDataSet(
    downloadLink="https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/TCGA.KIRC.sampleMap%2FKIRC_clinicalMatrix",
    filename="KIRC_clinicalMatrix",
    newFilename="KIRC_clinicalMatrix.txt",
    downloadDirectory="TCGA-KIRC",
    doUnzip=FALSE)

# Download, extract and rename the KIRC Curated survival data File
getDataSet(
    downloadLink="https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/survival%2FKIRC_survival.txt",
    filename="KIRC_survival.txt",
    newFilename="",
    downloadDirectory="TCGA-KIRC",
    doUnzip=FALSE)

# Clean up workspace
rm(getDataSet)
```

