---
title: "Differential Gene Expression — TCGA-KIRC (Tumour vs. Normal)"
subtitle: "Multi-Omics Integration for Survival Prediction in Clear Cell Renal Cell Carcinoma"
author: "Project Group 7 — Danilchyk, Shorzhina, Alves Radtke, Reinhart"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
    df_print: paged
params:
  data_dir: "TCGA-KIRC"
  lfc: 1.0             # |log2 fold-change| threshold for "significant"
  fdr: 0.05           # Benjamini-Hochberg FDR threshold
  min_detect_frac: 0.20  # keep genes expressed (>0) in at least this fraction of samples
  n_label: 15         # genes labelled on the volcano (each direction)
  n_heatmap: 50       # top genes shown in the heatmap
  seed: 42
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = knitr::is_html_output(),      # foldable code in HTML, hidden in the PDF report
  warning = FALSE, message = FALSE,
  cache = TRUE, cache.lazy = FALSE, autodep = TRUE,
  fig.width = 8, fig.height = 5, fig.align = "center", dpi = 120,
  out.width = "100%"
)
set.seed(params$seed)
```

# Introduction

This is **Step 2** of the pipeline (after the EDA notebook): A differential gene-expression (DGE) analysis contrasting **primary tumour vs. adjacent normal kidney** on the TCGA-KIRC RNA-seq layer. Its purpose is twofold:

1. **Biological validation** - confirm the cohort reproduces the well-established ccRCC transcriptional signature (e.g. hypoxia/HIF targets up, normal-tubule genes down). This grounds the later, survival-focused analyses in known tumour-vs-normal signal.
2. **Feature grounding** - the ranked gene list and the tumour-vs-normal signature feed interpretation of the MOFA factors and the survival models downstream.

The Xena RNA-seq matrix is **`log2(RSEM normalized_count + 1)`** - already normalised, continuous values. The correct tool for continuous log-expression is **`limma`** with an empirical-Bayes moderated *t*-test; we enable `trend = TRUE` (mean–variance trend, appropriate for log-count data) and `robust = TRUE` (robust to outlier genes).
```{r libraries}
suppressPackageStartupMessages({
  library(data.table)
  library(matrixStats)
  library(limma)          # moderated t-test for log-expression
  library(survival)       # univariate Cox bridge to survival
  library(pheatmap)       # heatmap of top DE genes
  library(ggrepel)        # volcano labels
  library(patchwork)
  library(RColorBrewer)
  library(clusterProfiler)  # GO over-representation
  library(org.Hs.eg.db)     # gene annotation
  library(fgsea)            # GSEA on ranked list
  library(msigdbr)          # Hallmark gene sets
  library(tidyverse)        # LAST (dplyr on top)
})

# clusterProfiler / AnnotationDbi export S4 generics (select, filter, rename, slice,
# count, ...) that can shadow dplyr's under knitr's cached package loading. Pinning them
# into the global environment (searched first) makes the notebook independent of load order.
select    <- dplyr::select
filter    <- dplyr::filter
rename    <- dplyr::rename
slice     <- dplyr::slice
count     <- dplyr::count
arrange   <- dplyr::arrange
mutate    <- dplyr::mutate
summarise <- dplyr::summarise
desc      <- dplyr::desc

DATA <- params$data_dir
path <- function(...) file.path(DATA, ...)
stopifnot(dir.exists(DATA))
type_cols <- c("Tumor" = "#D7301F", "Normal" = "#2171B5")
```

```{r helpers}
# Same barcode/loader helpers as the EDA notebook.
parse_barcode <- function(x) {
  parts <- str_split_fixed(x, "-", 4)
  code  <- substr(parts[, 4], 1, 2)
  type  <- dplyr::case_when(code == "01" ~ "Tumor",
                            code == "11" ~ "Normal",
                            code == "05" ~ "AddlPrimary",
                            TRUE ~ "Other")
  tibble(sample = x,
         patient = paste(parts[, 1], parts[, 2], parts[, 3], sep = "-"),
         type_code = code, sample_type = type)
}
load_matrix <- function(file) {
  dt <- fread(file, sep = "\t", header = TRUE, showProgress = FALSE)
  ids <- dt[[1]]; m <- as.matrix(dt[, -1]); rownames(m) <- ids
  storage.mode(m) <- "double"; m
}
```

# Data and experimental design

```{r load}
rna <- load_matrix(path("HiSeqV2.txt"))
bc  <- parse_barcode(colnames(rna))

# Two-group comparison: tumour vs solid-tissue normal (11). The tumour group is primary
# tumour (01) plus the single Additional-New-Primary (05) sample, which we treat as a
# tumour (it is malignant tissue). Its patient (TCGA-DV-A4W0) also has an 01 sample,
# so that patient contributes two tumours, so it's a negligible pseudoreplication (1 of 534).
bc$sample_type[bc$sample_type == "AddlPrimary"] <- "Tumor"
keep_samp   <- bc$sample_type %in% c("Tumor", "Normal")
rna         <- rna[, keep_samp]
group_info  <- bc[keep_samp, ]
group       <- factor(group_info$sample_type, levels = c("Normal", "Tumor"))

# Patient sex (used as a model covariate, see below).
pheno <- fread(path("KIRC_clinicalMatrix.txt"), sep = "\t", header = TRUE, na.strings = c("", "NA"))
sex   <- factor(setNames(pheno$gender, pheno$sampleID)[colnames(rna)])

knitr::kable(
  as.data.frame(table(group), stringsAsFactors = FALSE) |>
    setNames(c("group", "n_samples")),
  caption = "Samples entering the DGE analysis."
)
```

## Note: sex is a covariate, and the contrast is confounded with composition

Two design points that shape how the results must be read:

- **Sex is unbalanced between the groups** (below), and unadjusted this leaks sex-linked genes (XIST, Y-chromosome genes) into the "DE" list. We therefore fit **sex as a covariate** so the tumour-vs-normal contrast is adjusted for it. Sex is recorded for all samples, so no data is lost.
- **The contrast is confounded with cell-type composition.** Adjacent "normal" is almost pure kidney epithelium, whereas the tumour is malignant cells **plus infiltrating immune and stromal cells**. A substantial part of the "immune-response *up* / kidney-development *down*" signal is therefore **compositional** (immune cells are *present* in tumours; normal tubule cells are *absent*), not necessarily a per-cell transcriptional change. Fully separating intrinsic from compositional signal would require cell-type deconvolution or single-cell data; here we simply flag it so the enrichment is not over-interpreted.

```{r sex-balance}
knitr::kable(as.data.frame.matrix(table(group, sex)),
             caption = "Sex composition per group (motivates adjusting for sex).")
```

## Gene filtering

We drop genes that are effectively unexpressed (detected in fewer than
`r 100 * params$min_detect_frac` % of samples). Low-expression genes carry little signal and
inflate multiple-testing correction.

```{r gene-filter}
detect_frac <- rowMeans(rna > 0)
keep_gene   <- detect_frac >= params$min_detect_frac
expr        <- rna[keep_gene, ]
cat(sprintf("Genes: %d kept of %d (detected in >= %.0f%% of samples).\n",
            nrow(expr), nrow(rna), 100 * params$min_detect_frac))
```

# Differential expression (limma)

The model is `~ sex + group`; the coefficient of interest is the **tumour-vs-normal effect
adjusted for sex** (`groupTumor`).

```{r limma-fit}
design <- model.matrix(~ sex + group)     # adjust for sex; groupTumor = tumour effect
fit <- lmFit(expr, design)
fit <- eBayes(fit, trend = TRUE, robust = TRUE)

# Per-group means. `AveExpr` alone is misleading for on/off genes: e.g. SEMG2 is
# expressed in ~all normals (mean ~4.3) but silenced in the 534 tumours (~0.1), so its
# AveExpr (~0.6) looks like "low expression" when it is actually a clean marker. The
# per-group means keep the magnitude and direction honest.
mean_normal <- rowMeans(expr[, group == "Normal", drop = FALSE])
mean_tumor  <- rowMeans(expr[, group == "Tumor",  drop = FALSE])

res <- topTable(fit, coef = "groupTumor", number = Inf, sort.by = "P") |>
  rownames_to_column("gene") |>
  as_tibble() |>
  mutate(
    meanNormal = mean_normal[gene],
    meanTumor  = mean_tumor[gene],
    direction = case_when(
      adj.P.Val < params$fdr & logFC >=  params$lfc ~ "Up in tumour",
      adj.P.Val < params$fdr & logFC <= -params$lfc ~ "Down in tumour",
      TRUE ~ "n.s."
    )
  )

summary_tab <- res |> count(direction, name = "n_genes")
knitr::kable(summary_tab,
             caption = sprintf("DE gene counts (|log2FC| >= %.1f, FDR < %.2f).",
                               params$lfc, params$fdr))
```

### Significance is saturated: effect size does the real filtering

With 534 vs 72 samples, statistical power is enormous: essentially every gene reaches a tiny p-value (the smallest adjusted p is ≈ `r signif(min(res$adj.P.Val), 2)`), so the FDR cut alone would call a huge fraction of the transcriptome "significant". The biologically meaningful filter here is the **fold-change threshold**, not the p-value. The table below shows how the DE count shrinks as we tighten `|log2FC|`; we keep `|log2FC| ≥ `r params$lfc`` as the primary definition but report the stricter cuts for transparency.

```{r lfc-thresholds}
map_dfr(c(1, 1.5, 2), function(l) {
  tibble(`|log2FC| >=` = l,
         up_in_tumour   = sum(res$adj.P.Val < params$fdr & res$logFC >=  l),
         down_in_tumour = sum(res$adj.P.Val < params$fdr & res$logFC <= -l))
}) |>
  knitr::kable(caption = "DE counts at increasing fold-change thresholds (FDR < 0.05).")
```

```{r de-top-table}
p_fmt <- function(p) formatC(p, format = "e", digits = 2)   # avoid kable rounding p to 0
top_dir <- function(dir, n = 12) {
  res |> filter(direction == dir) |> slice_min(P.Value, n = n) |>
    transmute(gene, logFC = round(logFC, 2),
              meanNormal = round(meanNormal, 2), meanTumor = round(meanTumor, 2),
              t = round(t, 1), adj.P.Val = p_fmt(adj.P.Val), direction)
}
bind_rows(top_dir("Up in tumour"), top_dir("Down in tumour")) |>
  knitr::kable(caption = "Top 12 up- and top 12 down-regulated genes (by significance).")
```

# Visualisation

## Volcano plot

```{r volcano, fig.width=9, fig.height=6.5}
#markers_up   <- c("CA9", "NDUFA4L2", "VEGFA", "ANGPTL4", "EGLN3", "NNMT", "SLC6A3")
#markers_down <- c("UMOD", "SLC12A1", "CASR", "KCNJ1", "SLC34A1", "TMEM213")
#known <- intersect(c(markers_up, markers_down), res$gene)

lab <- bind_rows(
  res |> filter(direction == "Up in tumour")   |> slice_min(P.Value, n = params$n_label),
  res |> filter(direction == "Down in tumour") |> slice_min(P.Value, n = params$n_label),
  #res |> filter(gene %in% known)
) |> distinct(gene, .keep_all = TRUE)

ggplot(res, aes(logFC, -log10(P.Value))) +
  geom_point(aes(colour = direction), size = 0.7, alpha = 0.6) +
  scale_colour_manual(values = c("Up in tumour" = "#D7301F",
                                 "Down in tumour" = "#2171B5", "n.s." = "grey80")) +
  geom_vline(xintercept = c(-params$lfc, params$lfc), linetype = "dashed", colour = "grey50") +
  geom_hline(yintercept = -log10(params$fdr), linetype = "dashed", colour = "grey50") +
  geom_text_repel(data = lab, aes(label = gene), size = 2.8, max.overlaps = 30,
                  segment.size = 0.2, min.segment.length = 0) +
  labs(title = "Tumour vs. normal - volcano", x = "log2 fold-change (tumour / normal)",
       y = expression(-log[10]~italic(P)), colour = NULL) +
  theme_minimal(base_size = 12)
```

## MA plot and p-value histogram

```{r ma-pval, fig.width=10, fig.height=4}
p_ma <- ggplot(res, aes(AveExpr, logFC, colour = direction)) +
  geom_point(size = 0.6, alpha = 0.5) +
  scale_colour_manual(values = c("Up in tumour" = "#D7301F",
                                 "Down in tumour" = "#2171B5", "n.s." = "grey80"),
                      guide = "none") +
  geom_hline(yintercept = 0, colour = "grey40") +
  labs(title = "MA plot", x = "Average log2 expression", y = "log2 fold-change") +
  theme_minimal(base_size = 12)

p_hist <- ggplot(res, aes(P.Value)) +
  geom_histogram(boundary = 0, bins = 40, fill = "#7A0177", colour = "white") +
  labs(title = "P-value distribution", x = "p-value", y = "genes") +
  theme_minimal(base_size = 12)

p_ma | p_hist
```

A p-value histogram is roughly uniform under the null with a spike near zero; the MA plot shows symmetric, expression-independent fold-changes.

## Heatmap of top DE genes

We show the top up- and top down-regulated genes, so both arms of the signature are visible.

```{r heatmap, fig.width=9, fig.height=8}
half <- params$n_heatmap %/% 2
top_genes <- c(
  res |> filter(direction == "Up in tumour")   |> slice_min(adj.P.Val, n = half) |> pull(gene),
  res |> filter(direction == "Down in tumour") |> slice_min(adj.P.Val, n = half) |> pull(gene)
)

mat <- expr[top_genes, ]
mat_z <- t(scale(t(mat)))                       # z-score each gene across samples
ann_col <- data.frame(sample_type = group)
rownames(ann_col) <- colnames(mat_z)

pheatmap(mat_z,
         annotation_col = ann_col,
         annotation_colors = list(sample_type = type_cols),
         show_colnames = FALSE, show_rownames = TRUE, fontsize_row = 6,
         color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(50),
         breaks = seq(-3, 3, length.out = 51),
         clustering_method = "ward.D2",
         main = sprintf("Top %d DE genes (z-scored expression)", params$n_heatmap))
```

## Canonical marker boxplots (sanity check)

```{r marker-box, fig.width=10, fig.height=5}
show_markers <- intersect(c("CA9", "NDUFA4L2", "VEGFA", "UMOD", "SLC12A1", "CASR"),
                          rownames(expr))
mk_long <- as.data.frame(t(expr[show_markers, ])) |>
  rownames_to_column("sample") |>
  mutate(group = group) |>
  pivot_longer(all_of(show_markers), names_to = "gene", values_to = "expr") |>
  mutate(gene = factor(gene, levels = show_markers))

ggplot(mk_long, aes(group, expr, fill = group)) +
  geom_boxplot(outlier.size = 0.4) +
  scale_fill_manual(values = type_cols, guide = "none") +
  facet_wrap(~ gene, scales = "free_y", nrow = 2) +
  labs(title = "Canonical ccRCC markers (up) and normal-tubule markers (down)",
       x = NULL, y = "log2(count + 1)") +
  theme_minimal(base_size = 11)
```

## Quantifying the compositional confound: immune-infiltrate score

To turn the composition caveat into a measured quantity, we score each sample with a pan-immune marker panel (mean z-scored expression of leukocyte / antigen-presentation markers).

```{r immune-score, fig.width=6, fig.height=4.5}
immune_markers <- intersect(
  c("PTPRC", "CD3D", "CD3E", "CD8A", "CD2", "LCK", "CD68", "CD14", "CD163",
    "MS4A1", "CD19", "NKG7", "GZMB"),
  rownames(expr))
imm_z  <- t(scale(t(expr[immune_markers, ])))
imm_df <- tibble(sample = colnames(expr), group = group, immune_score = colMeans(imm_z))
wp <- wilcox.test(immune_score ~ group, data = imm_df)$p.value

ggplot(imm_df, aes(group, immune_score, fill = group)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_fill_manual(values = type_cols, guide = "none") +
  labs(title = "Immune-infiltrate score by group",
       subtitle = sprintf("%d leukocyte markers; Wilcoxon p = %s",
                          length(immune_markers), format.pval(wp, digits = 2)),
       x = NULL, y = "mean z(immune markers)") +
  theme_minimal(base_size = 12)
```

Tumours carry a much higher immune-marker load, so a large part of the immune enrichment
reflects **infiltrating leukocytes present in the tumour but absent from adjacent normal
tissue**, not a tumour-cell-intrinsic programme.

## External validation: overlap with a curated ccRCC signature

Beyond the qualitative marker check, we test the DE list against a **curated set of literature ccRCC markers** (hypoxia/HIF and metabolic genes up; nephron/tubule genes down): are they enriched in our DE set (hypergeometric test), and do they move in the expected direction?

```{r external-sig}
sig_up <- c("CA9","NDUFA4L2","VEGFA","ANGPTL4","EGLN3","NNMT","SLC6A3","VIM","CXCR4",
            "PLIN2","ENO2","SLC2A1","HILPDA","BNIP3","IGFBP3")
sig_dn <- c("UMOD","SLC12A1","CASR","KCNJ1","SLC34A1","TMEM213","ATP6V0A4","CLDN8",
            "FXYD4","NPHS2","SLC22A2","CALB1","AQP2","SLC12A3")
sig <- intersect(c(sig_up, sig_dn), res$gene)
de  <- res |> filter(direction != "n.s.") |> pull(gene)
ov  <- intersect(sig, de)
hp  <- phyper(length(ov) - 1, length(de), nrow(res) - length(de), length(sig),
              lower.tail = FALSE)
conc <- res |> filter(gene %in% sig) |>
  mutate(expected   = ifelse(gene %in% sig_up, "Up in tumour", "Down in tumour"),
         concordant = direction == expected)
cat(sprintf("Curated markers tested: %d | recovered as DE: %d | hypergeometric p = %.1e\n",
            length(sig), length(ov), hp))
cat(sprintf("Direction-concordant with literature: %d / %d\n",
            sum(conc$concordant), nrow(conc)))
```

The curated markers are recovered as DE with the correct direction — an independent confirmation that this is the genuine ccRCC programme, not a cohort-specific artefact.

# Robustness: matched tumour–normal pairs

Most normal samples come from the same patients as tumours. A **paired** analysis (blocking on patient) removes inter-patient variation and is a strong check that the signature is not driven by unmatched-cohort differences. We repeat limma on the matched subset with patient as a blocking factor and compare the gene lists.

```{r paired}
paired_patients <- group_info |>
  count(patient, sample_type) |>
  pivot_wider(names_from = sample_type, values_from = n, values_fill = 0) |>
  filter(Tumor >= 1, Normal >= 1) |> pull(patient)

pair_idx <- group_info$patient %in% paired_patients
# one tumour + one normal per patient (first of each) to keep a clean paired design
pair_tab <- group_info[pair_idx, ] |>
  group_by(patient, sample_type) |> slice_head(n = 1) |> ungroup()

expr_p  <- expr[, pair_tab$sample]
grp_p   <- factor(pair_tab$sample_type, levels = c("Normal", "Tumor"))
pat_p   <- factor(pair_tab$patient)

design_p <- model.matrix(~ pat_p + grp_p)
fit_p <- eBayes(lmFit(expr_p, design_p), trend = TRUE, robust = TRUE)
res_p <- topTable(fit_p, coef = "grp_pTumor", number = Inf) |>
  rownames_to_column("gene") |> as_tibble()

cat(sprintf("Matched pairs: %d patients (%d samples).\n",
            length(paired_patients), ncol(expr_p)))

de_unpaired <- res   |> filter(direction != "n.s.") |> pull(gene)
de_paired   <- res_p |> filter(adj.P.Val < params$fdr, abs(logFC) >= params$lfc) |> pull(gene)
cat(sprintf("DE genes unpaired: %d | paired: %d | overlap: %d (%.1f%% of paired)\n",
            length(de_unpaired), length(de_paired),
            length(intersect(de_unpaired, de_paired)),
            100 * length(intersect(de_unpaired, de_paired)) / max(1, length(de_paired))))
```

```{r paired-concordance, fig.width=6, fig.height=5}
comp <- inner_join(res |> select(gene, lfc_unpaired = logFC),
                   res_p |> select(gene, lfc_paired = logFC), by = "gene")
ggplot(comp, aes(lfc_unpaired, lfc_paired)) +
  geom_point(size = 0.5, alpha = 0.3, colour = "#4292C6") +
  geom_abline(slope = 1, intercept = 0, colour = "grey40", linetype = "dashed") +
  labs(title = "log2FC concordance: unpaired vs. paired",
       x = "log2FC (all samples)", y = "log2FC (matched pairs)") +
  theme_minimal(base_size = 12)
```

# Biological interpretation (enrichment)

We annotate the signature with two complementary approaches: **GO over-representation** on the significant up/down gene sets, and **GSEA** on the full ranked list (Hallmark gene sets).

> **Read with the composition caveat.** Several of the strongest up-in-tumour terms
> (allograft rejection, interferon response, inflammatory response, T-cell activation) are
> **immune-infiltrate** signals - the leukocytes quantified in the immune-score figure above
> - not necessarily tumour-cell-intrinsic transcription. Hypoxia, glycolysis, E2F/G2M
> proliferation and the OXPHOS/fatty-acid-metabolism *down* shift are more plausibly
> tumour-intrinsic. We flag which leading-edge genes are immune-lineage below.

## GO over-representation (Biological Process)

```{r go-ora, fig.width=10, fig.height=7}
universe <- res$gene
run_go <- function(genes, title) {
  eg <- enrichGO(gene = genes, universe = universe, OrgDb = org.Hs.eg.db,
                 keyType = "SYMBOL", ont = "BP", pAdjustMethod = "BH",
                 pvalueCutoff = 0.05, qvalueCutoff = 0.10, readable = FALSE)
  if (is.null(eg) || nrow(as.data.frame(eg)) == 0) return(NULL)
  # Collapse semantically redundant GO terms for a cleaner, non-repetitive figure.
  eg <- clusterProfiler::simplify(eg, cutoff = 0.7, by = "p.adjust", select_fun = min)
  as.data.frame(eg) |> slice_head(n = 12) |> mutate(set = title)
}

up_genes   <- res |> filter(direction == "Up in tumour")   |> pull(gene)
down_genes <- res |> filter(direction == "Down in tumour") |> pull(gene)
go_df <- bind_rows(run_go(up_genes, "Up in tumour"),
                   run_go(down_genes, "Down in tumour"))

if (!is.null(go_df) && nrow(go_df) > 0) {
  go_df |>
    group_by(set) |> slice_head(n = 10) |> ungroup() |>
    mutate(Description = fct_reorder(Description, -log10(p.adjust))) |>
    ggplot(aes(-log10(p.adjust), Description, fill = set)) +
    geom_col() +
    scale_fill_manual(values = c("Up in tumour" = "#D7301F", "Down in tumour" = "#2171B5"),
                      guide = "none") +
    facet_wrap(~ set, scales = "free", ncol = 1) +
    labs(title = "GO Biological Process enrichment", x = expression(-log[10]~adj.~italic(P)),
         y = NULL) +
    theme_minimal(base_size = 10)
} else {
  cat("No significant GO terms returned.\n")
}
```

## GSEA on Hallmark gene sets

```{r gsea, fig.width=9, fig.height=6}
ranks <- res |> filter(is.finite(t)) |> arrange(desc(t)) |>
  distinct(gene, .keep_all = TRUE) |> select(gene, t) |> deframe()

# msigdbr >= 10 renamed `category` to `collection`; support both.
hm <- tryCatch(msigdbr(species = "Homo sapiens", collection = "H"),
               error = function(e) msigdbr(species = "Homo sapiens", category = "H"))
hallmark <- split(hm$gene_symbol, hm$gs_name)

fg <- fgsea(pathways = hallmark, stats = ranks, minSize = 10, maxSize = 500) |>
  as_tibble() |> arrange(padj)

fg |> slice_head(n = 15) |>
  transmute(pathway = str_replace_all(str_remove(pathway, "HALLMARK_"), "_", " "),
            NES = round(NES, 2), padj = signif(padj, 3), size) |>
  knitr::kable(caption = "Top Hallmark gene sets by adjusted p (GSEA).")

fg |> filter(padj < 0.05) |> slice_max(abs(NES), n = 15) |>
  mutate(pathway = fct_reorder(str_remove(pathway, "HALLMARK_"), NES)) |>
  ggplot(aes(NES, pathway, fill = NES > 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "#D7301F", "FALSE" = "#2171B5"),
                    labels = c("TRUE" = "up in tumour", "FALSE" = "down in tumour"),
                    name = NULL) +
  labs(title = "Hallmark GSEA (FDR < 0.05)", x = "Normalised enrichment score", y = NULL) +
  theme_minimal(base_size = 11)
```

**Leading-edge genes** - the core genes driving the strongest enrichments - make the
signature concrete and give candidate features for the downstream models. The last column
counts how many of the shown leading-edge genes are **immune-lineage** markers (a proxy for
how compositional, vs. tumour-intrinsic, each enrichment is).

The immune-lineage reference is **not hand-picked**: we take the union of the mature-leukocyte
cell-type signatures from the **`HAY_BONE_MARROW`** hematopoietic single-cell atlas (Hay et
al., *Exp. Hematol.* 2018), as distributed in **MSigDB C8** and loaded here via `msigdbr`
(v`r as.character(packageVersion("msigdbr"))`). We explicitly select the differentiated
immune populations and exclude progenitor, erythroid, platelet and stromal clusters. As a
sanity check we print the panel size and its immune-fraction enrichment for a known immune
vs. a known metabolic pathway.

```{r gsea-leading-edge}
# Versioned, sourced immune panel: mature leukocyte populations from the HAY bone-marrow
# single-cell reference (MSigDB C8, Hay et al. 2018). Progenitors/erythroid/platelet/stromal
# clusters are deliberately excluded so the panel is leukocyte-specific.
hay_immune_sets <- c(
  "HAY_BONE_MARROW_CD8_T_CELL", "HAY_BONE_MARROW_NAIVE_T_CELL",
  "HAY_BONE_MARROW_FOLLICULAR_B_CELL", "HAY_BONE_MARROW_PRO_B",
  "HAY_BONE_MARROW_NK_CELLS", "HAY_BONE_MARROW_MONOCYTE",
  "HAY_BONE_MARROW_DENDRITIC_CELL", "HAY_BONE_MARROW_PRE_DENDRITIC",
  "HAY_BONE_MARROW_NEUTROPHIL", "HAY_BONE_MARROW_IMMATURE_NEUTROPHIL",
  "HAY_BONE_MARROW_EOSINOPHIL", "HAY_BONE_MARROW_PLASMA_CELL")
c8 <- tryCatch(msigdbr(species = "Homo sapiens", collection = "C8"),
               error = function(e) msigdbr(species = "Homo sapiens", category = "C8"))
immune_lineage <- c8 |> filter(gs_name %in% hay_immune_sets) |>
  distinct(gene_symbol) |> pull(gene_symbol)

is_immune <- function(g) mean(g %in% immune_lineage)  # immune fraction of a gene vector
cat(sprintf("Immune-lineage panel (HAY C8): %d genes (%.1f%% of tested).\n",
            length(immune_lineage), 100 * length(immune_lineage) / nrow(res)))
cat(sprintf("Immune fraction of leading edge — ALLOGRAFT_REJECTION: %.2f | OXIDATIVE_PHOSPHORYLATION: %.2f\n",
            is_immune(head(fg$leadingEdge[[which(fg$pathway == "HALLMARK_ALLOGRAFT_REJECTION")]], 8)),
            is_immune(head(fg$leadingEdge[[which(fg$pathway == "HALLMARK_OXIDATIVE_PHOSPHORYLATION")]], 8))))

fg |> slice_head(n = 6) |>
  mutate(pathway = str_replace_all(str_remove(pathway, "HALLMARK_"), "_", " "),
         NES = round(NES, 2),
         le  = map(leadingEdge, ~ head(.x, 8)),
         `leading-edge genes` = map_chr(le, ~ paste(.x, collapse = ", ")),
         `immune / shown` = map_chr(le, ~ sprintf("%d / %d", sum(.x %in% immune_lineage), length(.x)))) |>
  select(pathway, NES, `leading-edge genes`, `immune / shown`) |>
  knitr::kable(caption = "Leading-edge genes for the top Hallmark pathways (immune-lineage count, from the HAY C8 panel, flags compositional signal).")
```

# Bridge to the survival question: are DE genes prognostic?

This DGE step **validates the data** — it confirms the expected tumour-vs-normal biology. But the project's actual question is survival *within tumours*, and a strong tumour-vs-normal gene is **not** automatically prognostic. We fit, for every DE gene (overall survival, tumours only), both a **univariate** Cox model and one **adjusted for stage and grade** — the dominant ccRCC prognosticators — to see how much apparent prognostic signal is independent of them. We report **hazard ratios** (effect size), not just significance counts, because with ~520 tumours univariate p-values are again saturated.

```{r cox-bridge, fig.width=10, fig.height=4.5}
surv <- fread(path("KIRC_survival.txt"),
              sep = "\t", header = TRUE, na.strings = c("", "NA"))
tum <- colnames(expr)[group == "Tumor"]
cl  <- tibble(sample = tum,
              OS = surv$OS[match(tum, surv$sample)], OS.time = surv$OS.time[match(tum, surv$sample)],
              stage = pheno$pathologic_stage[match(tum, pheno$sampleID)],
              grade = pheno$neoplasm_histologic_grade[match(tum, pheno$sampleID)]) |>
  mutate(stage = factor(str_replace(stage, "\\[Discrepancy\\]", NA_character_),
                        levels = c("Stage I","Stage II","Stage III","Stage IV")),
         grade = factor(na_if(grade, "GX"), levels = c("G1","G2","G3","G4")))
keepc <- with(cl, !is.na(OS) & !is.na(OS.time) & OS.time > 0 & !is.na(stage) & !is.na(grade))
tumK  <- cl$sample[keepc]
Sy    <- Surv(cl$OS.time[keepc], cl$OS[keepc])
stage <- droplevels(cl$stage[keepc]); grade <- droplevels(cl$grade[keepc])

de_genes <- res |> filter(direction != "n.s.") |> pull(gene)
X <- t(expr[de_genes, tumK])                       # tumours x DE genes

# Univariate Cox and a stage+grade-adjusted Cox for every DE gene.
cox_stat <- function(j) {
  x  <- as.numeric(scale(X[, j]))
  f1 <- tryCatch(coxph(Sy ~ x), error = function(e) NULL)
  f2 <- tryCatch(coxph(Sy ~ x + stage + grade), error = function(e) NULL)
  c(if (is.null(f1)) NA else summary(f1)$coefficients[1, "coef"],
    if (is.null(f1)) NA else summary(f1)$coefficients[1, "Pr(>|z|)"],
    if (is.null(f2)) NA else summary(f2)$coefficients["x", "Pr(>|z|)"])
}
cx  <- t(vapply(seq_len(ncol(X)), cox_stat, numeric(3)))
cox <- tibble(gene = colnames(X), cox_coef = cx[, 1], cox_p = cx[, 2], adj_p = cx[, 3]) |>
  mutate(HR = exp(cox_coef), cox_fdr = p.adjust(cox_p, "BH"), adj_fdr = p.adjust(adj_p, "BH")) |>
  inner_join(res |> select(gene, logFC, direction), by = "gene")

n_uni  <- sum(cox$cox_fdr < 0.05, na.rm = TRUE)
n_adj  <- sum(cox$adj_fdr < 0.05, na.rm = TRUE)
n_both <- sum(cox$cox_fdr < 0.05 & cox$adj_fdr < 0.05, na.rm = TRUE)
r_signed <- cor(cox$logFC, cox$cox_coef, use = "complete.obs")
cat(sprintf("Adjusted-Cox cohort: %d tumours (complete OS + stage + grade).\n", length(tumK)))
cat(sprintf("Prognostic DE genes  univariate: %d (%.0f%%)  |  stage+grade-adjusted: %d (%.0f%%)\n",
            n_uni, 100*n_uni/nrow(cox), n_adj, 100*n_adj/nrow(cox)))
cat(sprintf("  -> %.0f%% of univariate hits survive stage/grade adjustment;\n the rest were largely stage/grade proxies.\n",
            100 * n_both / max(1, n_uni)))
cat(sprintf("Prognostic-gene hazard ratios (per SD): median %.2f, IQR [%.2f, %.2f].\n",
            median(cox$HR[cox$cox_fdr < 0.05], na.rm = TRUE),
            quantile(cox$HR[cox$cox_fdr < 0.05], .25, na.rm = TRUE),
            quantile(cox$HR[cox$cox_fdr < 0.05], .75, na.rm = TRUE)))
cat(sprintf("cor(tumour-vs-normal log2FC, Cox log-HR) = %.2f\n  (weak positive: up-in-tumour leans to worse survival)\n", r_signed))

# Signed: does UP in tumour mean WORSE survival? (log2FC vs Cox log-HR)
p_signed <- ggplot(cox, aes(logFC, cox_coef, colour = direction)) +
  geom_point(size = 0.5, alpha = 0.35) +
  scale_colour_manual(values = c("Up in tumour"="#D7301F","Down in tumour"="#2171B5"), guide = "none") +
  geom_hline(yintercept = 0, colour = "grey50") + geom_vline(xintercept = 0, colour = "grey50") +
  labs(title = "Direction vs. prognosis", x = "log2FC (tumour vs. normal)",
       y = "Cox log-HR (OS)") + theme_minimal(base_size = 11)

top_prog <- cox |> filter(cox_fdr < 0.05) |> slice_min(cox_p, n = 15) |>
  mutate(gene = fct_reorder(gene, -log10(cox_p)))
p_top <- ggplot(top_prog, aes(-log10(cox_p), gene, fill = HR > 1)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE"="#D7301F","FALSE"="#2171B5"),
                    labels = c("TRUE"="worse surv.","FALSE"="better surv."), name = NULL) +
  labs(title = "Most prognostic DE genes", x = expression(-log[10]~Cox~italic(P)), y = NULL) +
  theme_minimal(base_size = 10)

p_signed | p_top
```

Takeaways for the survival notebook: **(1)** a large share of DE genes look univariately
prognostic, but only **`r round(100*n_both/max(1,n_uni))`%** survive adjustment for stage and
grade - much of the raw "prognostic" signal is a **stage/grade proxy**; **(2)** the strongest
*independent* signal is a **cell-cycle / proliferation** programme (e.g. CDCA3, UBE2C, CCNF,
TROAP) associated with worse survival - a non-compositional, tumour-intrinsic effect; and
**(3)** the tumour programme leans *weakly* toward worse prognosis (signed log2FC-vs-log-HR
r ≈ `r round(r_signed, 2)`; left panel), but the association is loose - tumour-vs-normal
magnitude explains only a few percent of prognostic variance. So the survival models must
**learn survival-relevant features directly**, not inherit them from this contrast. (Hence we
use the DE list only as an *unsupervised* expression filter, never selecting features with
survival labels — avoiding information leakage into cross-validation.)

# Export results

The full ranked table is written to disk for downstream feature selection, MOFA-factor
interpretation, and the survival models.

The table also carries the univariate-Cox prognostic columns (OS, tumours only) so the
downstream notebooks can use them directly.

```{r export}
out_file <- "DGE_tumor_vs_normal_limma.csv"
res |>
  left_join(cox |> select(gene, cox_HR = HR, cox_p, cox_fdr, cox_adj_fdr = adj_fdr),
            by = "gene") |>
  transmute(gene, logFC, meanNormal, meanTumor, AveExpr, t, P.Value, adj.P.Val, direction,
            cox_HR, cox_p, cox_fdr, cox_adj_fdr) |>
  arrange(adj.P.Val) |>
  write_csv(out_file)
```

# Summary

Comparing **534 tumour (533 primary + 1 additional-primary) vs. 72 adjacent normal** kidney samples (18,684 genes tested, sex-adjusted `limma`), the analysis produced the following results.

- **A large, balanced, biologically correct signature.** `r sum(res$direction != "n.s.")` genes are differentially expressed at |log2FC| ≥ `r params$lfc` and FDR < `r params$fdr` - `r sum(res$direction == "Up in tumour")` up and `r sum(res$direction == "Down in tumour")` down in tumour.

- **Significance is saturated: effect size does the filtering.** With 605 samples the smallest adjusted p ≈ `r signif(min(res$adj.P.Val), 2)`; essentially the whole transcriptome is "significant", so the |log2FC| threshold, not the FDR, defines the gene set (DE shrinks 4,943 → 2,968 → 1,861 as the cut tightens 1.0 → 1.5 → 2.0). The most extreme fold-changes are genuine **on/off genes**;

- **Robust to design.** The matched-pair analysis (72 patients, patient-blocked) recovers **91.5 %** of the DE genes with tightly concordant fold-changes - the signature is not an artefact of comparing unmatched cohorts.

- **Pathways, read with a *measured* composition caveat.** Hallmark GSEA is up for immune/inflammatory (allograft rejection, IFN-γ/α, IL6-JAK-STAT3, inflammatory response), proliferation (E2F targets, G2M) and hypoxia/glycolysis/EMT, and down for oxidative phosphorylation, fatty-acid and bile-acid metabolism. But an immune-infiltrate score is far higher in tumours (**Wilcoxon p < 2e-16**), so much of the immune signal reflects **infiltrating leukocytes** (present in tumour, absent in normal), not tumour-cell-intrinsic transcription. The hypoxia/glycolysis-up, OXPHOS/fatty-acid-down metabolic switch is the tumour-intrinsic core.

- **DE is not prognostic, and much apparent prognosis is stage-driven.** On the `r length(tumK)` tumours with complete outcome data, `r n_uni` DE genes (`r round(100*n_uni/nrow(cox))` %) are univariately prognostic for overall survival, but only `r n_adj` (`r round(100*n_adj/nrow(cox))` %) remain after adjusting for **stage and grade** - ≈ `r round(100*n_both/max(1,n_uni))` % of univariate hits survive, the rest being stage/grade proxies (prognostic HRs are modest: median `r round(median(cox$HR[cox$cox_fdr<0.05], na.rm=TRUE), 2)` per SD). Tumour-vs-normal fold-change correlates only weakly with prognosis (signed log2FC-vs-log-HR r ≈ `r round(r_signed, 2)`): the tumour programme leans slightly toward worse survival but explains little prognostic variance. The clearest *independent* signal is a **proliferation programme** (CDCA3, UBE2C, CCNF, TROAP, KIF18B, CHFR - worse survival). The survival models must therefore learn prognostic features directly; the DE list is used only as an **unsupervised** expression filter (no survival labels), avoiding leakage.

```{r session-info}
sessionInfo()
# Removes all objects in the current workspace
rm(list = ls(all.names = TRUE))
# Frees up memory
gc() 
```
